%\documentclass[twocolumn]{amsart}
\documentclass[journal]{IEEEtran}
\usepackage[cmex10]{amsmath}
\usepackage{tikz, amssymb,listings}

\begin{document}

%\bibliographystyle{plain}
\bibliographystyle{IEEEtran}

\title{MAC address table size design}

\author{Morten~Jagd~Christensen, Ph.D.~\IEEEmembership{Senior Member,~IEEE}% <-this % stops a space

John~Doe~Smith
%\thanks{M. Christensen is with Cobham, (e-mail: morten.jagd.christensen @ cobham.com).}% <-this % stops a space
}

\markboth{Journal of \LaTeX\ Class Files,~Vol.~6, No.~1, January~2007}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}

\maketitle

\begin{abstract}
A central feature of an Ethernet switch is the size and organization of its MAC address table. The table 
is often populated by means of hashing MAC addresses and other parameters. Often manufacturers 
use the size of this table as a marketing parameter, but the size of the table is misleading when it comes to
performance.
In this article we give a brief description of Ethernet switching followed by discussions of hashing in the context of 
Ethernet switching. A statistical model for hash collisions and how they relate to performance is presented.
Finally some specific implementations of hashing functions are described and these are used 
in simulations.  In conclusion we discuss how size constraints and performance of the hashing functions 
can lead to design descisions with regards to MAC address table implementations.
\end{abstract}


\input{introduction}
\input{binsnbuckets}
\input{hashcolls}
\input{hashfuncs}
\input{simulations}


\section{Conclusion}
\IEEEPARstart{W}{hen}
considering ,...


\section*{Acknowledgment}
\IEEEPARstart{T}{he}
authors would like to thank Professor Christian Kornerup Hansen, Eastern Washington University, for his
suggestions to the content and organization of this paper.


\bibliography{Master}



\end{document}
